Race Science in Mainstream Psychology, 1960-2010

### 2025-02-03 paper/paper.qmd
ready to resubmit?

Histories of scientific racism in the twentieth century have often emphasized the Pioneer Fund and the parascholarly journal *Mankind Quarterly* (MQ) [
@BarkanRetreatScientificRacism1992;
@MehlerFoundationFascismNew1989;
@WinstonScienceServiceFar1998;
@TuckerFundingScientificRacism2002;
@SchafferScientificRacismAgain2007;
@SainiSuperiorReturnRace2019;
@WinstonScientificRacismNorth2020;
@AdamsMisappropriationBiologicalAnthropology2021;
@SainiDraperMillionsPhilanthropic2022].  Pioneer was formed in 1937, during the waning years of the eugenics movement in North America [@BarkanRetreatScientificRacism1992; @TuckerFundingScientificRacism2002], with the aim of "support[ing] academic research and the \'dissemination of information, into the 'problem of heredity and eugenics' and 'the problems of race betterment'" [@MehlerFoundationFascismNew1989 p.21, quoting Laughlin, unbalanced quotation marks in Mehler].  MQ was founded in 1960 by biologist R. Ruggles Gates (1882-1962), psychologist Henry Garrett (1894-1973), and non-academic anthropologist G. Robert Gayre (self-styled as "Gayre of Gayre and Nigg"; 1907-1996).  Gates' professional status had risen and fallen with eugenics and the explicit scientific racism of the 1920s, and by the end of the Second World War he was thoroughly marginalized [@WinstonScienceServiceFar1998].  In contrast, Garrett had been president of the American Psychological Association in 1946 and chair of Psychology at Columbia from 1941 to 1955 [@WinstonScienceServiceFar1998].

@@

>>>>

Histories of scientific racism in the twentieth century have often emphasized the Pioneer Fund and the parascholarly journal *Mankind Quarterly* (MQ) [
@BarkanRetreatScientificRacism1992;
@MehlerFoundationFascismNew1989;
@WinstonScienceServiceFar1998;
@TuckerFundingScientificRacism2002;
@SchafferScientificRacismAgain2007;
@SainiSuperiorReturnRace2019;
@WinstonScientificRacismNorth2020;
@AdamsMisappropriationBiologicalAnthropology2021;
@SainiDraperMillionsPhilanthropic2022].  Pioneer was formed in 1937, during the waning years of the eugenics movement in North America [@BarkanRetreatScientificRacism1992; @TuckerFundingScientificRacism2002], with the aim of "support[ing] academic research and the \'dissemination of information, into the 'problem of heredity and eugenics' and 'the problems of race betterment'" [@MehlerFoundationFascismNew1989 p.21, quoting Laughlin, unbalanced quotation marks in Mehler].  MQ was founded in 1960 by biologist R. Ruggles Gates (1882-1962), psychologist Henry Garrett (1894-1973), and non-academic anthropologist G. Robert Gayre (self-styled as "Gayre of Gayre and Nigg"; 1907-1996).  Gates' professional status had risen and fallen with eugenics and the explicit scientific racism of the 1920s, and by the end of the Second World War he was thoroughly marginalized [@WinstonScienceServiceFar1998].  In contrast, Garrett had been president of the American Psychological Association in 1946 and chair of Psychology at Columbia from 1941 to 1955 [@WinstonScienceServiceFar1998].  [@ShukmanRaceScienceInc2024 report that the Pioneer Fund has been effectively renamed the Human Diversity Foundation (HDF).]

@@

(Late in the publication cycle, it occurred to us that evolutionary psychology is often associated with race science in the same way as behavior genetics. Because no evolutionary psychology journals were included in the corpus, we are unable to examine this association here. However, it would be a worthwhile topic for future research.)
